---
title: "DelayedTensor"
author:
- name: Koki Tsuyuzaki
  affiliation: Laboratory for Bioinformatics Research,
    RIKEN Center for Biosystems Dynamics Research
- name: Itoshi Nikaido
  affiliation: Laboratory for Bioinformatics Research,
    RIKEN Center for Biosystems Dynamics Research
  email: k.t.the-answer@hotmail.co.jp
graphics: no
package: DelayedTensor
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{DelayedTensor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Authors**: `r packageDescription("DelayedTensor")[["Author"]] `<br />
**Last modified:** `r file.info("DelayedTensor.Rmd")$mtime`<br />
**Compiled**: `r date()`

# Introduction
## Concept of DelayedTensor

# Tensor arithmetic operations
## Setting

## Unfold operations
unfold(), k_unfold(), matvec(), rs_unfold(), cs_unfold(), ttl()

## Fold operations
fold(), k_fold(), unmatvec(), rs_fold(), cs_fold(), ttm()

## Vectorization
vec()

## Norm operations
fnorm(), innerProd()

## Diagonal operations/Diagonal Tensor
diag(), DelayedDiagonalArray()

## Mode-wise operations
modeSum(), modeMean()

## Tensor product operations
hadamard(), hadamard_list(), kronecker(), kronecker_list(), khatri_rao,
khatri_rao_list()

## Utilities
list_rep(), modebind_list(), rbind_list(), cbind_list()

# Tensor Decomposition
hosvd(), cp(), tucker(), mpca(), pvd()

# Einsum operation
einsum()

# Session information {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```